---
title: "Choosing k: Rank Selection and Diagnostics"
author: "Michael Totty"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Choosing k: Rank Selection and Diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  eval = FALSE
)
```

*Note: This vignette contains code that downloads datasets and runs
computationally intensive analyses. Set `eval = TRUE` or run chunks
interactively to execute.*

# Introduction

The number of NMF programs (k) is the most important parameter in any NMF analysis. Choosing k too low merges distinct biological programs; choosing k too high splits coherent programs into noisy fragments. NMFscape provides several complementary diagnostics to guide this decision.

```{r load-packages, message=FALSE}
library(NMFscape)
library(SingleCellExperiment)
library(scuttle)
library(scater)
library(scRNAseq)
library(ggplot2)

set.seed(427)
```

## Prepare data

We use the Zeisel brain dataset with ~3,000 cells across 7 major cell types.

```{r prepare-data}
sce <- ZeiselBrainData()
sce <- sce[!grepl("^mt-", rownames(sce)), ]
sce <- logNormCounts(sce)
```

# Choosing the Right Distribution

Before selecting k, consider which loss function is appropriate for your data. NMFscape supports Gaussian (MSE), generalized Poisson, negative binomial, and other distributions via RcppML.

`selectDistribution()` compares multiple loss functions and recommends the best one based on BIC:

```{r select-distribution}
dist_result <- selectDistribution(sce, k = 10, verbose = FALSE)

# Best distribution
dist_result$best

# Full comparison
dist_result$comparison
```

For log-normalized data, MSE is typically the best choice. For raw counts, Poisson or negative binomial may be more appropriate.

# Cross-Validation for Rank Selection

`selectRank()` uses held-out cross-validation: a fraction of matrix entries are masked during training, then reconstruction error is evaluated on the masked entries. The optimal k minimizes test error.

```{r cross-validation}
cv_results <- selectRank(sce, k = seq(2, 20, by = 2),
                         test_fraction = 0.2, verbose = FALSE)
cv_results
```

Visualize the results as an elbow plot. The dashed red line marks the rank with lowest test error:

```{r plot-cv, fig.width=6, fig.height=4}
plotRankSelection(cv_results)
```

For more robust estimates, run multiple replicates:

```{r cv-replicates}
cv_reps <- selectRank(sce, k = seq(2, 20, by = 2),
                      test_fraction = 0.2, n_reps = 3,
                      seed = 42, verbose = FALSE)

# Error bars show standard deviation across replicates
plotRankSelection(cv_reps)
```

# Consensus Stability

Another approach is to check how stable the factorization is across random initializations. `consensusNMF()` runs NMF multiple times and measures agreement via the cophenetic correlation coefficient. Higher values indicate more stable, reproducible factorizations.

```{r consensus, message=FALSE}
# Test stability at k = 5, 10, 15
cophenetic_scores <- vapply(c(5, 10, 15), function(ki) {
    sce_tmp <- consensusNMF(sce, k = ki, n_runs = 10, verbose = FALSE)
    metadata(sce_tmp)$cNMF_consensus$cophenetic
}, numeric(1))

data.frame(k = c(5, 10, 15), cophenetic = round(cophenetic_scores, 4))
```

A cophenetic score close to 1 indicates high stability. A sharp drop suggests the factorization is becoming unstable at that rank.

# Assessing Against Known Labels

When cell type annotations are available, `assessNMF()` evaluates how well the NMF embedding separates known groups using clustering metrics (ARI, NMI, silhouette) and cross-validated classification accuracy.

```{r assess-labels}
sce <- runNMFscape(sce, k = 10, verbose = FALSE)

assessment <- assessNMF(sce, label_col = "level1class", min_class_size = 5)

# Key metrics
data.frame(
    metric = c("ARI", "NMI", "Silhouette", "KNN Accuracy"),
    value = round(c(
        assessment$metrics$ari,
        assessment$metrics$nmi,
        assessment$metrics$silhouette,
        assessment$metrics$accuracy_knn
    ), 3)
)
```

# Model Evaluation

`evaluateNMF()` computes the reconstruction loss for a fitted model --- how well does the factorization approximate the original data?

```{r evaluate}
loss <- evaluateNMF(sce)
cat("Reconstruction loss (MSE):", round(loss, 4), "\n")
```

Lower loss indicates better fit, but beware of overfitting at very high k.

# Practical Recommendations

| Diagnostic | What it tells you | When to use |
|---|---|---|
| `selectRank()` | Optimal k by held-out reconstruction error | Always --- this is the primary diagnostic |
| `selectDistribution()` | Best loss function for your data type | Before running NMF, especially with count data |
| `consensusNMF()` | Factorization stability at a given k | When cross-validation is ambiguous |
| `assessNMF()` | Agreement with known biology | When cell type labels are available |
| `evaluateNMF()` | Reconstruction quality | Comparing models with different parameters |

In practice, we recommend:

1. Start with `selectRank()` to get a cross-validation curve
2. Pick k near the elbow (where test error stops decreasing)
3. Validate with `assessNMF()` if labels are available
4. Check `consensusNMF()` if the elbow is ambiguous

For most single-cell brain datasets, k between 10--30 works well. Larger, more heterogeneous datasets may need higher k.

# Session Information

```{r session-info, eval = TRUE}
sessionInfo()
```
