---
title: "Deep NMF: Hierarchical Gene Programs"
author: "Michael Totty"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Deep NMF: Hierarchical Gene Programs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  fig.retina = 1,
  dpi = 96
)
```

# Introduction

Brain cell types are organized hierarchically: neurons and glia form the broadest division, which subdivides into excitatory/inhibitory neurons, astrocytes, oligodendrocytes, etc. Deep (multi-layer) NMF captures this hierarchy naturally --- the first layer learns fine-grained gene programs, and subsequent layers group them into meta-programs that correspond to broader biological categories.

NMFscape's `runDeepNMF()` wraps RcppML's FactorNet graph DSL to make hierarchical factorization as simple as specifying a vector of ranks.

```{r load-packages, message=FALSE}
library(NMFscape)
library(SingleCellExperiment)
library(ggplot2)
library(patchwork)

set.seed(427)
```

# Data Preparation

To make the hierarchy visible we simulate it directly: five meta-programs,
each built from four finer programs, each of which drives twenty genes. Cells
belong to one meta-program and express its four fine programs. This is the
structure `runDeepNMF()` is designed to recover, and it keeps the vignette
fast and self-contained. Everything below works the same way on a real
`SingleCellExperiment`.

```{r prepare-data}
n_meta <- 5
fine_per_meta <- 4
n_fine <- n_meta * fine_per_meta
genes_per_fine <- 20
n_genes <- n_fine * genes_per_fine
n_cells <- 500

# gene loadings: each fine program owns a block of genes
W_true <- matrix(0.05, n_genes, n_fine)
for (j in seq_len(n_fine)) {
    rows <- ((j - 1) * genes_per_fine + 1):(j * genes_per_fine)
    W_true[rows, j] <- runif(genes_per_fine, 1, 3)
}

# cell loadings: each cell activates the fine programs of its meta-program
meta_of_fine <- rep(seq_len(n_meta), each = fine_per_meta)
cell_meta <- sample(seq_len(n_meta), n_cells, replace = TRUE)
H_true <- matrix(0.05, n_fine, n_cells)
for (i in seq_len(n_cells)) {
    fine <- which(meta_of_fine == cell_meta[i])
    H_true[fine, i] <- runif(length(fine), 1, 3)
}

mat <- W_true %*% H_true
mat <- mat * exp(rnorm(length(mat), 0, 0.15))
mat <- log1p(mat / max(mat) * 200)
rownames(mat) <- sprintf("Gene_%04d", seq_len(n_genes))
colnames(mat) <- sprintf("Cell_%03d", seq_len(n_cells))

sce <- SingleCellExperiment(list(logcounts = mat))
sce$celltype <- paste0("Type_", cell_meta)
sce
```

# Running Deep NMF

Specify a vector of ranks to define the hierarchy. Here, 20 fine-grained programs are compressed into 5 meta-programs:

```{r run-deep}
sce <- runDeepNMF(sce, k = c(20, 5), seed = 1, verbose = FALSE)
```

The results include:

- **Effective basis** (genes x 5): maps original features directly to meta-programs
- **Per-layer details**: the full W, H, and d for each layer
- **Cell embedding** (cells x 5): the meta-program coefficients for each cell

```{r check-results}
# Effective basis maps genes to 5 meta-programs
eff_basis <- getBasis(sce, "DeepNMF")
dim(eff_basis)

# Per-layer details
layers <- metadata(sce)$DeepNMF_layers
cat("Layer 1 W:", dim(layers$layer_1$W), "(genes x 20 programs)\n")
cat("Layer 2 W:", dim(layers$layer_2$W), "(cells x 5 meta-programs)\n")
```

# Exploring the Hierarchy

## Meta-program gene signatures

The effective basis tells us which genes define each meta-program. Since this is computed by propagating through all layers, it maps original genes directly to the final meta-programs:

```{r meta-signatures}
top_meta <- getTopFeatures(sce, name = "DeepNMF", n = 10)
top_meta
```

## Meta-programs vs cell types

```{r meta-dotplot, fig.width=6, fig.height=5}
plotProgramDots(sce, group = "celltype", nmf_name = "DeepNMF")
```

Because we know the ground truth here, we can check the recovery directly.
Assigning each cell to its strongest meta-program should reproduce the
simulated grouping:

```{r meta-recovery}
embedding <- reducedDim(sce, "DeepNMF")
table(truth = sce$celltype,
      recovered = paste0("M", apply(embedding, 1, which.max)))
```

On real data each meta-program typically corresponds to a broad biological
category (e.g. neurons, glia, immune cells) rather than a single cell type.

# Comparing to Single-Layer NMF

How does deep NMF compare to standard NMF at the same final rank?

```{r compare-standard}
# Standard NMF with k=5 (same final dimensionality)
sce <- runNMFscape(sce, k = 5, name = "NMF_k5", seed = 1, verbose = FALSE)

# Standard NMF with k=20 (same first-layer dimensionality)
sce <- runNMFscape(sce, k = 20, name = "NMF_k20", seed = 1, verbose = FALSE)
```

```{r compare-dotplots, fig.width=14, fig.height=5}
p1 <- plotProgramDots(sce, group = "celltype", nmf_name = "NMF_k5") +
    ggtitle("Standard NMF (k=5)")
p2 <- plotProgramDots(sce, group = "celltype", nmf_name = "DeepNMF") +
    ggtitle("Deep NMF (20 -> 5)")
p1 + p2
```

Deep NMF often produces cleaner meta-programs because the first layer captures fine-grained variation, and the second layer finds natural groupings among those programs rather than trying to directly compress from the full gene space.

# When to Use Deep NMF

Deep NMF is most useful when:

- Your data has a clear hierarchical structure (e.g., brain cell types)
- You want both fine-grained programs and interpretable meta-programs
- Standard NMF at low k produces mixed or ambiguous programs
- You need a compact embedding but don't want to lose biological detail

For datasets without clear hierarchy, standard `runNMFscape()` with a single k is simpler and often sufficient.

## More than two layers

`k` may have any length; the embedding is always `k[length(k)]`-dimensional
regardless of depth, because the basis and coefficients are propagated through
every layer.

```{r three-layers}
sce <- runDeepNMF(sce, k = c(20, 10, 5), name = "DeepNMF3",
                  seed = 1, verbose = FALSE)
dim(getBasis(sce, "DeepNMF3"))
dim(reducedDim(sce, "DeepNMF3"))
```

Note that deep results are stored as a `factor_net_result` rather than an S4
`nmf`, so `evaluateNMF()` and `predictNMF()` do not apply to them; the fitted
loss is available directly:

```{r deep-loss}
metadata(sce)$DeepNMF_model$total_loss
```

# Session Information

```{r session-info}
sessionInfo()
```
