---
title: "Deep NMF: Hierarchical Gene Programs"
author: "Michael Totty"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Deep NMF: Hierarchical Gene Programs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  eval = FALSE
)
```

*Note: This vignette contains code that downloads datasets and runs
computationally intensive analyses. Set `eval = TRUE` or run chunks
interactively to execute.*

# Introduction

Brain cell types are organized hierarchically: neurons and glia form the broadest division, which subdivides into excitatory/inhibitory neurons, astrocytes, oligodendrocytes, etc. Deep (multi-layer) NMF captures this hierarchy naturally --- the first layer learns fine-grained gene programs, and subsequent layers group them into meta-programs that correspond to broader biological categories.

NMFscape's `runDeepNMF()` wraps RcppML's FactorNet graph DSL to make hierarchical factorization as simple as specifying a vector of ranks.

```{r load-packages, message=FALSE}
library(NMFscape)
library(SingleCellExperiment)
library(scuttle)
library(scater)
library(scRNAseq)
library(ggplot2)
library(patchwork)

set.seed(427)
```

# Data Preparation

```{r prepare-data}
sce <- ZeiselBrainData()
sce <- sce[!grepl("^mt-", rownames(sce)), ]
sce <- logNormCounts(sce)
sce <- runPCA(sce, ncomponents = 20)
sce <- runUMAP(sce, dimred = "PCA")
```

# Running Deep NMF

Specify a vector of ranks to define the hierarchy. Here, 20 fine-grained programs are compressed into 5 meta-programs:

```{r run-deep}
sce <- runDeepNMF(sce, k = c(20, 5), verbose = FALSE)
```

The results include:

- **Effective basis** (genes x 5): maps original features directly to meta-programs
- **Per-layer details**: the full W, H, and d for each layer
- **Cell embedding** (cells x 5): the meta-program coefficients for each cell

```{r check-results}
# Effective basis maps genes to 5 meta-programs
eff_basis <- getBasis(sce, "DeepNMF")
dim(eff_basis)

# Per-layer details
layers <- metadata(sce)$DeepNMF_layers
cat("Layer 1 W:", dim(layers$layer_1$W), "(genes x 20 programs)\n")
cat("Layer 2 W:", dim(layers$layer_2$W), "(cells x 5 meta-programs)\n")
```

# Exploring the Hierarchy

## Meta-program gene signatures

The effective basis tells us which genes define each meta-program. Since this is computed by propagating through all layers, it maps original genes directly to the final meta-programs:

```{r meta-signatures}
top_meta <- getTopFeatures(sce, name = "DeepNMF", n = 10)
top_meta
```

## Meta-programs vs cell types

```{r meta-dotplot, fig.width=6, fig.height=5}
plotProgramDots(sce, group = "level1class", nmf_name = "DeepNMF")
```

With k=5 meta-programs and ~7 cell types, each meta-program typically corresponds to a broad biological category (e.g., neurons, glia, immune cells).

# Comparing to Single-Layer NMF

How does deep NMF compare to standard NMF at the same final rank?

```{r compare-standard}
# Standard NMF with k=5 (same final dimensionality)
sce <- runNMFscape(sce, k = 5, name = "NMF_k5", verbose = FALSE)

# Standard NMF with k=20 (same first-layer dimensionality)
sce <- runNMFscape(sce, k = 20, name = "NMF_k20", verbose = FALSE)
```

```{r compare-dotplots, fig.width=14, fig.height=5}
p1 <- plotProgramDots(sce, group = "level1class", nmf_name = "NMF_k5") +
    ggtitle("Standard NMF (k=5)")
p2 <- plotProgramDots(sce, group = "level1class", nmf_name = "DeepNMF") +
    ggtitle("Deep NMF (20 -> 5)")
p1 + p2
```

Deep NMF often produces cleaner meta-programs because the first layer captures fine-grained variation, and the second layer finds natural groupings among those programs rather than trying to directly compress from the full gene space.

# When to Use Deep NMF

Deep NMF is most useful when:

- Your data has a clear hierarchical structure (e.g., brain cell types)
- You want both fine-grained programs and interpretable meta-programs
- Standard NMF at low k produces mixed or ambiguous programs
- You need a compact embedding but don't want to lose biological detail

For datasets without clear hierarchy, standard `runNMFscape()` with a single k is simpler and often sufficient.

# Session Information

```{r session-info, eval = TRUE}
sessionInfo()
```
