---
title: "Getting Started with NMFscape"
author: "Michael Totty"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Getting Started with NMFscape}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 5,
  fig.retina = 1,
  dpi = 96
)
```

# Introduction

NMFscape provides fast, scalable non-negative matrix factorization (NMF) for single-cell and spatial transcriptomic data. NMF decomposes gene expression data into biologically interpretable gene expression programs (also called "factors" or "metagenes") and their weighted patterns across cells.

## What is NMF?

Non-negative matrix factorization decomposes a gene expression matrix **X** (genes x cells) into two lower-rank matrices:

- **W** (genes x k programs): Gene loadings - which genes contribute to each program
- **H** (k programs x cells): Cell loadings - how much each cell uses each program

The key constraint is that all values in W and H are non-negative (>=0), which enforces a "parts-based" representation where programs combine additively. This is particularly suited for biological data where gene expression counts are inherently non-negative.

## Why NMF for single-cell analysis?

- **Gene program decomposition**: Identifies gene modules that work together in biological processes
- **Cell type discovery**: Programs often correspond to cell type-specific signatures, cell states, or biological pathways
- **Biologically interpretable**: Gene programs represent co-expressed gene modules which can be used downstream for functional enrichment or pathway analysis

NMFscape uses [RcppML](https://github.com/zdebruine/RcppML) for high-performance computation and integrates seamlessly with Bioconductor workflows, storing results directly in `SingleCellExperiment` and `SpatialExperiment` objects.

```{r load-packages, message=FALSE}
library(NMFscape)
library(SingleCellExperiment)
library(scuttle)
library(scater)
library(scRNAseq)
library(ggplot2)
library(patchwork)

set.seed(427)
```

# Basic NMF Workflow

## Load and prepare data

We'll use the Zeisel brain dataset (Zeisel et al. 2015, *Science*) which contains ~3,000 cells from the mouse brain with diverse cell types including neurons, astrocytes, oligodendrocytes, and immune cells. We remove mitochondrial genes as they can dominate gene programs and typically reflect technical noise rather than biological signals.

```{r prepare-data}
sce <- ZeiselBrainData()

# Remove mitochondrial genes
sce <- sce[!grepl("^mt-", rownames(sce)), ]

# Log-normalize counts
sce <- logNormCounts(sce)

# Generate PCA and UMAP for visualization
sce <- runPCA(sce, ncomponents = 20)
sce <- runUMAP(sce, dimred = "PCA")
```

## Run NMF

Here we run NMF with k=10 programs. Choosing k is important and should be informed by cross-validation --- see the `vignette("choosing-k")` for a detailed guide using `selectRank()`.

NMFscape supports multiple loss functions via the `distribution` parameter (e.g., `"mse"`, `"gp"`, `"nb"`), or `"auto"` to automatically select the best distribution for your data.

```{r run-nmf}
sce <- runNMFscape(sce, k = 10, verbose = FALSE)

# NMF coefficients are stored in reducedDims
nmf_coords <- reducedDim(sce, "NMF")
dim(nmf_coords)

# The raw RcppML model is stored for downstream use
model <- getModel(sce)
model
```

# Visualizing NMF Results

## Plot NMF programs on UMAP

A key use of NMF is finding gene programs associated with specific cell types and biological states. Cell type-specific programs can be identified by plotting program weights in UMAP space.

```{r visualize-programs, fig.width=12, fig.height=6}
p1 <- plotReducedDim(sce, dimred = "UMAP", colour_by = "level1class",
                     point_size = 0.8) +
  ggtitle("Zeisel Brain Cell Types")

p2 <- vizUMAP(sce, program = 1) +
  ggtitle("NMF Program 1")

p1 + p2
```

## Finding cell type-specific programs

Dot plots summarize program usage across annotated cell types.

```{r dot-plots, fig.width=6, fig.height=5}
plotProgramDots(sce, group = "level1class")
```

## Examine top weighted genes per program

A heatmap of gene weights from the basis matrix (W) shows which genes define each program.

```{r gene-weights, fig.width=4, fig.height=8}
plotProgramHeatmap(sce, n_genes = 10, cluster_rows = FALSE)
```

We can verify that NMF programs are finding cell type-specific markers by comparing to expression across cell types:

```{r gene-expression, fig.width=4, fig.height=8}
top_genes <- getTopFeatures(sce, n = 10)

# Top 5 unique genes across all programs
top5_genes <- unique(unlist(lapply(top_genes, function(x) x[seq_len(5)])))

scater::plotGroupedHeatmap(sce, features = top5_genes,
                           group = "level1class",
                           assay = "logcounts",
                           scale = TRUE, center = TRUE)
```

# Finding Differentially Expressed Programs

## Differential program usage

`FindAllDEPs()` performs 1-vs-all comparisons for each cell type to identify programs that are significantly enriched or depleted.

```{r dep-analysis}
deps <- FindAllDEPs(sce, cell_type_col = "level1class")

# View results for pyramidal CA1 neurons
head(deps[["pyramidal CA1"]])
```

## Visualizing differential program usage

Volcano plots show the relationship between fold enrichment and statistical significance. Note that p-values from this analysis are liberal and should be interpreted as a ranking rather than strict significance thresholds.

```{r dep-volcano, fig.width=10, fig.height=8}
plotDEPsVolcano(deps, fold_threshold = 1.5, pval_threshold = 0.05)
```

A summary heatmap of differential program usage across all cell types:

```{r dep-heatmap, fig.width=3, fig.height=4}
plotDEPsHeatmap(deps, logfc_threshold = 2, star_size = 14)
```

# Next Steps

NMFscape provides a full suite of tools for NMF-based analysis:

- **Choosing k**: Cross-validation, distribution selection, and consensus stability for optimal rank selection. See `vignette("choosing-k")`.
- **Cell type programs**: Deeper biological interpretation with label-guided refinement. See `vignette("cell-type-programs")`.
- **Multi-modal NMF**: Joint factorization of RNA + chromatin accessibility data. See `vignette("multi-modal")`.
- **Transfer learning**: Project new data onto a reference NMF model with batch correction. See `vignette("transfer-learning")`.

# Session Information

```{r session-info}
sessionInfo()
```
