---
title: "Guided NMF and Batch Correction"
author: "Michael Totty"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Guided NMF and Batch Correction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  fig.retina = 1,
  dpi = 96
)
```

# Introduction

NMFscape has four different ways to push cell labels or a nuisance covariate
into a factorization, and they are easy to confuse. This vignette runs all
four on one simulated dataset that has both a known cell type signal and a
known batch effect, scores them on the same two axes, and reports what
actually happened -- including where a method did nothing useful.

| Function | When guidance is applied | Mechanism |
|---|---|---|
| `refineNMF()` | After fitting | Wraps `RcppML::refine()`; shifts an existing embedding toward class centroids, optionally batch-aware |
| `runGuidedNMF(mode = "enrich")` | During fitting | Adds a class-centroid target for `H` with a positive `target_lambda`, attracting the embedding toward the labels |
| `runConditionedNMF()` | During fitting | Wraps `RcppML::factor_condition()`; concatenates a covariate design matrix to `H` so the covariate is explained by dedicated columns rather than by the programs |

The two axes we care about are always the same:

1. **Does the embedding still separate cell types?** (we want yes)
2. **Does the embedding still separate batches?** (we want no)

```{r load-packages, message=FALSE}
library(NMFscape)
library(SingleCellExperiment)
library(ggplot2)

set.seed(1001)
```

# A simulation with a known batch effect

Each cell gets a cell type program (one of three, on genes 1--150) and a batch
program (one of two, on genes 151--250), plus noise. The batch program is
deliberately given a much larger amplitude than the cell type program, which
is the usual situation in real data: the technical axis is the loudest thing
in the matrix.

```{r simulate}
n_cells <- 450
n_genes <- 300
n_type <- 3
n_batch <- 2

celltype <- rep(paste0("Type", seq_len(n_type)), length.out = n_cells)
batch <- rep(paste0("Batch", seq_len(n_batch)), length.out = n_cells)

# cell type basis and coefficients (modest amplitude)
w_type <- matrix(0.2, n_genes, n_type)
for (i in seq_len(n_type)) {
  w_type[((i - 1) * 50 + 1):(i * 50), i] <- runif(50, 1.2, 1.9)
}
h_type <- matrix(0.2, n_type, n_cells)
for (j in seq_len(n_cells)) {
  h_type[match(celltype[j], sort(unique(celltype))), j] <- runif(1, 1, 1.5)
}

# batch basis and coefficients (large amplitude)
w_batch <- matrix(0.1, n_genes, n_batch)
for (i in seq_len(n_batch)) {
  w_batch[(150 + (i - 1) * 50 + 1):(150 + i * 50), i] <- runif(50, 3, 4.8)
}
h_batch <- matrix(0.1, n_batch, n_cells)
for (j in seq_len(n_cells)) {
  h_batch[match(batch[j], sort(unique(batch))), j] <- runif(1, 1, 1.5)
}

counts_mat <- w_type %*% h_type + w_batch %*% h_batch +
  matrix(abs(rnorm(n_genes * n_cells, 0, 2.5)), n_genes, n_cells)
rownames(counts_mat) <- sprintf("Gene_%03d", seq_len(n_genes))
colnames(counts_mat) <- sprintf("Cell_%03d", seq_len(n_cells))

sce <- SingleCellExperiment(list(logcounts = counts_mat))
sce$celltype <- celltype
sce$batch <- batch

k <- 6
sce
```

## Scoring both axes

`assessNMF()` returns clustering metrics (ARI, NMI, silhouette) and
cross-validated classifier accuracy for a stored embedding. We call it twice
per method -- once against `celltype`, once against `batch` -- and keep four
numbers.

The distinction between the two kinds of metric matters here:

* **ARI** asks whether unsupervised clustering of the embedding *lands on*
  the labels. It reflects which axis dominates the geometry.
* **Classifier accuracy** asks whether the labels are *recoverable at all*
  from the embedding. It reflects whether the information is still present.

A method can move the first without touching the second.

```{r score-helper}
scoreEmbedding <- function(x, name) {
  by_type <- suppressMessages(
    assessNMF(x, name = name, label_col = "celltype", min_class_size = 10)
  )
  by_batch <- suppressMessages(
    assessNMF(x, name = name, label_col = "batch", min_class_size = 10)
  )
  data.frame(
    method = name,
    type_ari = round(by_type$metrics$ari, 3),
    type_acc = round(by_type$metrics$accuracy_knn, 3),
    batch_ari = round(by_batch$metrics$ari, 3),
    batch_acc = round(by_batch$metrics$accuracy_knn, 3)
  )
}
```

# Baseline: unguided NMF

```{r baseline}
sce <- runNMFscape(sce, k = k, name = "NMF", seed = 42, verbose = FALSE)
scoreEmbedding(sce, "NMF")
```

This is the problem in one row. Cell type is perfectly *recoverable*
(`type_acc = 1`), but clustering the embedding recovers **batch**, not cell
type (`batch_ari = 1`, `type_ari = 0.16`). Any downstream clustering step run
on this embedding will find batches and call them cell types.

# refineNMF(): post hoc correction

`refineNMF()` takes the fitted model and shifts each cell's embedding toward
its class centroid. It does not refit from the data, so it is cheap, and
`batch_col` makes the shift batch-aware.

```{r refine}
sce <- refineNMF(sce, label_col = "celltype", name = "NMF",
                 refined_name = "Refined", batch_col = "batch",
                 verbose = FALSE)
scoreEmbedding(sce, "Refined")
```

Cell type becomes the dominant clustering axis (`type_ari` 0.16 to 1.00) and
batch stops being one (`batch_ari` 1.00 to ~0). Note that `batch_acc` stays at
1.00: the batch information is still sitting in the embedding, it is simply no
longer the direction that clustering picks up.

# runGuidedNMF(mode = "enrich"): guidance during fitting

`runGuidedNMF()` builds a class-centroid target for `H` with
`RcppML::compute_target()` and refits with a positive `target_lambda`. Because
the target has to be computed from *some* embedding, the function fits twice:
an unguided base model, then the guided model. Set `init_name` to keep the
base fit as well, which makes the guided-versus-unguided comparison free.

Guidance strength matters, and the response is not gradual. Here is a sweep:

```{r enrich-sweep}
enrich_scores <- list()
for (s in c(1, 2, 5, 10)) {
  nm <- paste0("Enrich_", s)
  sce <- runGuidedNMF(sce, k = k, label_col = "celltype", mode = "enrich",
                      strength = s, name = nm, seed = 42, verbose = FALSE)
  enrich_scores[[nm]] <- cbind(strength = s, scoreEmbedding(sce, nm))
}
do.call(rbind, enrich_scores)
```

Below strength 5 the guidance is simply too weak to compete with the batch
program and the result is bit-for-bit uninteresting -- identical scores to the
unguided fit. At strength 5 it flips completely and stays there. This is worth
knowing: if guidance appears to do nothing, the strength is too low, not the
method wrong.

The default `whiten = TRUE` also matters. `compute_target()` applies
OAS-shrinkage ZCA whitening when building the centroids, and turning it off
loses the effect entirely at this strength:

```{r whiten}
sce <- runGuidedNMF(sce, k = k, label_col = "celltype", mode = "enrich",
                    strength = 5, whiten = FALSE, name = "Enrich_nowhiten",
                    seed = 42, verbose = FALSE)
scoreEmbedding(sce, "Enrich_nowhiten")
```

We keep strength 5 with whitening as the representative enrichment result:

```{r enrich-final}
sce <- runGuidedNMF(sce, k = k, label_col = "celltype", mode = "enrich",
                    strength = 5, name = "Enriched", init_name = "Unguided",
                    seed = 42, verbose = FALSE)
scoreEmbedding(sce, "Enriched")
```

## Semi-supervised use: labelling only some cells

`compute_target()` gives `NA` cells a zero target column, so you can label the
cells you are confident about and pass `NA` for the rest. No cells are dropped.

```{r partial-labels}
set.seed(123)
partial <- sce$celltype
partial[sample(ncol(sce), round(ncol(sce) * 0.6))] <- NA
sce$partial_type <- partial

sce <- runGuidedNMF(sce, k = k, label_col = "partial_type", mode = "enrich",
                    strength = 5, name = "PartialEnrich", seed = 42,
                    verbose = FALSE)

metadata(sce)$PartialEnrich_guidance$n_guided
nrow(reducedDim(sce, "PartialEnrich"))
scoreEmbedding(sce, "PartialEnrich")
```

With 40% of the cells labelled, strength 5 -- which flipped the embedding
completely with full labels -- does not flip it at all. Cell type ARI stays
near the unguided baseline.

The instinct is to turn the strength up, and that runs into the second thing
worth knowing: **a zero target column is not an exemption from the penalty, it
is a target of zero.** Unlabelled cells are pulled toward the origin rather
than left to float, and the pull grows with `strength`. We can measure it by
comparing the coefficient mass of unlabelled to labelled cells:

```{r partial-shrinkage}
labelled <- !is.na(sce$partial_type)
strengths <- c(1, 5, 20, 50)

mass_ratio <- vapply(strengths, function(s) {
  nm <- paste0("Shrink_", s)
  fit <- runGuidedNMF(sce, k = k, label_col = "partial_type", mode = "enrich",
                      strength = s, name = nm, seed = 42, verbose = FALSE)
  emb <- reducedDim(fit, nm)
  mean(rowSums(emb[!labelled, ])) / mean(rowSums(emb[labelled, ]))
}, numeric(1))

data.frame(strength = strengths, unlabelled_over_labelled = round(mass_ratio, 3))
```

At strength 1 the unlabelled cells are untouched. As the strength climbs they
lose coefficient mass, and in a dataset with less competing signal than this
one the collapse is sharper still -- on a batch-free version of this
simulation the ratio fell to roughly one third by strength 20.

The practical rule: semi-supervised guidance is real, but it is a low-strength
tool. Keep `strength` modest when a large fraction of cells is unlabelled, and
check that the unlabelled cells have not collapsed toward the origin. If you
need a high strength to overcome a dominant nuisance axis, as you do on this
simulation, partial labels will not get you there -- label more cells, or use
`refineNMF()`, which shifts only the cells that have labels and leaves the
rest exactly where they were.

# Why `mode = "remove"` is disabled

`runGuidedNMF()` also accepts a *negative* `target_lambda`, which triggers
RcppML's eigenvalue-projected adversarial update: instead of attracting `H`
toward the target it suppresses structure correlated with it. In principle
that is exactly what batch removal should be, and it is the one operation here
that would genuinely *delete* a covariate rather than merely demote it.

In practice it destroys the factorization, so `mode = "remove"` errors rather
than returning a degenerate result:

```{r remove-refused, error = TRUE}
runGuidedNMF(sce, k = k, label_col = "batch", mode = "remove",
             strength = 0.5, verbose = FALSE)
```

Across strengths from 0.5 to 10 on both simulated and real data, the embedding
collapsed onto a single program -- every cell taking its maximum on the same
factor, with most programs exactly zero -- and agreement with cell type fell to
ARI 0.000 at every strength tested. It is a cliff with no usable window rather
than a dial. If you need a covariate to be genuinely unrecoverable, neither
this nor anything else in the package delivers that today; see the summary
table below for what the remaining methods do and do not achieve.

# runConditionedNMF(): factoring the covariate out

`runConditionedNMF()` uses a completely different mechanism. It one-hot encodes
the covariate and concatenates it to `H` via `RcppML::factor_condition()`, so
the model has explicit columns available to explain batch and, in principle,
does not need to spend NMF programs on it.

```{r conditioned}
sce <- runConditionedNMF(sce, k = k, condition_col = "batch",
                         name = "Conditioned", seed = 42, verbose = FALSE)
scoreEmbedding(sce, "Conditioned")
```

On this simulation conditioning did nothing measurable. The embedding is not
identical to the unguided fit -- the numbers move -- but every metric rounds to
the same value, and batch is still the dominant clustering axis. The batch
programs here are strong, dense, and sit on their own gene block, so NMF has
every incentive to keep modelling them regardless of the extra conditioning
columns. Conditioning is a soft constraint, not a projection.

# Side by side

```{r summary-table}
summary_tab <- do.call(rbind, lapply(
  c("NMF", "Refined", "Enriched", "Conditioned"),
  function(nm) scoreEmbedding(sce, nm)
))
summary_tab$method <- c("unguided NMF", "refineNMF", "guided: enrich",
                        "runConditionedNMF")
summary_tab
```

```{r summary-plot, fig.height = 4}
plot_df <- data.frame(
  method = rep(summary_tab$method, 4),
  metric = rep(c("cell type ARI", "cell type accuracy",
                 "batch ARI", "batch accuracy"), each = nrow(summary_tab)),
  value = c(summary_tab$type_ari, summary_tab$type_acc,
            summary_tab$batch_ari, summary_tab$batch_acc)
)
plot_df$method <- factor(plot_df$method, levels = rev(summary_tab$method))
plot_df$metric <- factor(plot_df$metric, levels = c(
  "cell type ARI", "cell type accuracy", "batch ARI", "batch accuracy"))

ggplot(plot_df, aes(x = value, y = method, color = metric)) +
  geom_point(size = 3) +
  facet_wrap(~ metric, nrow = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
  labs(x = "score", y = NULL,
       title = "Higher is better on the left two panels, lower on the right two") +
  theme_bw() +
  theme(legend.position = "none")
```

What the numbers say:

* **`refineNMF()` and `runGuidedNMF(mode = "enrich")` are equally effective
  here.** Both take cell type ARI from 0.16 to 1.00 and drop batch ARI to
  approximately 0. Neither erases batch: `batch_acc` stays at 1.00 for both.
  They reorganise which axis dominates the geometry, they do not remove the
  covariate.
* **`refineNMF()` is much cheaper.** It reuses the existing fit;
  `runGuidedNMF()` fits twice by construction. On this data that is a fraction
  of a second either way, but on a real dataset it is the difference between
  one factorization and three.

* **`runConditionedNMF()` had no measurable effect here.** That is a statement
  about this simulation, not a general verdict -- conditioning is better suited
  to covariates that are diffuse across many genes, or continuous covariates
  such as depth or percent mitochondrial reads, where a couple of extra `H`
  columns can absorb the trend. Against a strong, block-structured batch
  effect it does not compete.

# Which should I use when

| Situation | Reach for | Why |
|---|---|---|
| You have trustworthy labels and an already-fitted model | `refineNMF()` | Post hoc, no refit, as effective as guided fitting on this benchmark |
| You want labels to shape the programs themselves, not just the embedding | `runGuidedNMF(mode = "enrich")` | Guidance enters the optimization, so `W` is learned under the constraint |
| Only some cells are labelled | `runGuidedNMF(mode = "enrich")` at low `strength`, or `refineNMF()` | `NA` cells get a zero target column: at low strength that is effectively no guidance, but at high strength it shrinks them toward the origin |
| The covariate must be genuinely unrecoverable (e.g. donor identity) | none of these | Adversarial removal is disabled; all remaining methods demote the covariate rather than deleting it |
| Continuous or diffuse nuisance variation (depth, %mito) | `runConditionedNMF()` | Dedicated `H` columns can absorb a trend without a target |
| You just want to know whether you have a problem | `assessNMF()` twice | Score the embedding against biology and against the covariate before choosing anything |

Two closing caveats.

Guidance is only available through the single-layer `RcppML::nmf()` path that
`runGuidedNMF()` uses. The per-factor configuration objects behind
`runDeepNMF()` and the other FactorNet recipes accept target arguments but
ignore them, so there is no guided deep NMF and no `W`-side (marker gene)
anchoring.

And guided methods make the labels true by construction. If you guide with
cluster assignments and then evaluate against those same assignments, you will
measure your own input. Validate on something the guidance did not see.

# Session info

```{r session-info}
sessionInfo()
```
