---
title: "Multi-Modal NMF Analysis"
author: "Michael Totty"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Multi-Modal NMF Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  eval = FALSE
)
```

*Note: This vignette contains code that downloads datasets and runs
computationally intensive analyses. Set `eval = TRUE` or run chunks
interactively to execute.*

# Introduction

Multi-modal single-cell assays (10x Multiome, CITE-seq, SHARE-seq) measure multiple data types from the same cells. NMFscape's `runMultiModalNMF()` performs joint NMF across modalities, learning a **shared cell embedding** while allowing each modality to have its own **feature loadings**. This finds programs that are consistent across data types.

This vignette demonstrates the approach using simulated multi-modal data from the Zeisel brain dataset. In practice, you would use real multiome data (e.g., RNA + ATAC from 10x Multiome).

```{r load-packages, message=FALSE}
library(NMFscape)
library(SingleCellExperiment)
library(scuttle)
library(scater)
library(scRNAseq)
library(ggplot2)

set.seed(427)
```

# Setting Up Multi-Modal Data

We simulate a second modality by selecting a subset of genes and adding noise, stored as an `altExp` in the SCE. In real data, this would be chromatin accessibility peaks, surface protein levels, etc.

```{r prepare-data}
sce <- ZeiselBrainData()
sce <- sce[!grepl("^mt-", rownames(sce)), ]
sce <- logNormCounts(sce)
sce <- runPCA(sce, ncomponents = 20)
sce <- runUMAP(sce, dimred = "PCA")

# Simulate a second modality (e.g., "ATAC-like")
# Use a random subset of genes with added noise
set.seed(123)
atac_genes <- sample(rownames(sce), 500)
atac_mat <- as.matrix(assay(sce, "logcounts")[atac_genes, ])
atac_mat <- atac_mat + abs(matrix(rnorm(length(atac_mat), sd = 0.5),
                                  nrow = nrow(atac_mat)))

atac_sce <- SingleCellExperiment(list(counts = atac_mat))
rownames(atac_sce) <- paste0("peak_", seq_len(nrow(atac_mat)))
colnames(atac_sce) <- colnames(sce)
altExp(sce, "ATAC") <- atac_sce
```

The SCE now has two modalities:

- **Main assay** ("logcounts"): the full gene set
- **altExp "ATAC"**: 500 simulated peaks

Both share the same cells.

# Running Multi-Modal NMF

`runMultiModalNMF()` accepts modalities via the `alt_exps` parameter. Use `"main"` to refer to the primary SCE assay:

```{r run-multimodal}
sce <- runMultiModalNMF(sce, k = 8,
                        alt_exps = c("main", "ATAC"),
                        main_assay = "logcounts",
                        alt_assay = "counts",
                        modality_names = c("RNA", "ATAC"),
                        verbose = FALSE)
```

# Shared Cell Embedding

The shared H matrix captures cell identity using information from both modalities:

```{r shared-embedding, fig.width=12, fig.height=5}
p1 <- plotReducedDim(sce, dimred = "UMAP", colour_by = "level1class",
                     point_size = 0.8) +
    ggtitle("Cell Types (UMAP)")

# Also run single-modality NMF for comparison
sce <- runNMFscape(sce, k = 8, name = "NMF_RNA", verbose = FALSE)

p1
```

# Per-Modality Feature Loadings

Each modality gets its own basis matrix. Use the `modality` parameter in `getBasis()`:

```{r modality-basis}
rna_basis <- getBasis(sce, "MultiNMF", modality = "RNA")
cat("RNA basis:", dim(rna_basis), "\n")

atac_basis <- getBasis(sce, "MultiNMF", modality = "ATAC")
cat("ATAC basis:", dim(atac_basis), "\n")
```

## Top features per modality

```{r top-features}
# Top RNA genes per program
top_rna <- getTopFeatures(sce, name = "MultiNMF", n = 5, modality = "RNA")
cat("Top RNA genes for Program 1:\n")
top_rna[[1]]

# Top peaks per program
top_atac <- getTopFeatures(sce, name = "MultiNMF", n = 5, modality = "ATAC")
cat("\nTop peaks for Program 1:\n")
top_atac[[1]]
```

## Program usage across cell types

The shared embedding can be visualized the same way as standard NMF:

```{r dotplot, fig.width=6, fig.height=5}
plotProgramDots(sce, group = "level1class", nmf_name = "MultiNMF")
```

# When to Use Multi-Modal NMF

Real use cases for `runMultiModalNMF()`:

- **10x Multiome**: Joint RNA + ATAC-seq from the same nuclei. Store ATAC peaks as `altExp`.
- **CITE-seq**: RNA + surface protein (ADT). Protein panel as `altExp`.
- **SHARE-seq**: RNA + ATAC with combinatorial indexing.
- **Spatial + scRNA-seq**: When cells are matched across modalities.

The key advantage over running NMF separately on each modality is that the shared embedding enforces consistency --- programs must explain variation in *both* data types, filtering out modality-specific technical noise.

**Note:** This vignette used simulated data for illustration. With real multiome data, the multi-modal embedding typically outperforms single-modality NMF for cell type separation and reveals regulatory relationships between gene expression and chromatin state.

# Session Information

```{r session-info, eval = TRUE}
sessionInfo()
```
