---
title: "Multi-Modal NMF Analysis"
author: "Michael Totty"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Multi-Modal NMF Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6
)
```

# Introduction

Multi-modal single-cell assays (10x Multiome, CITE-seq, SHARE-seq) measure multiple data types from the same cells. NMFscape's `runMultiModalNMF()` performs joint NMF across modalities, learning a **shared cell embedding** while allowing each modality to have its own **feature loadings**. This finds programs that are consistent across data types.

This vignette demonstrates the approach on simulated multiome-like data. In
practice you would use real multiome data (e.g. RNA + ATAC from 10x Multiome).

```{r load-packages, message=FALSE}
library(NMFscape)
library(SingleCellExperiment)
library(ggplot2)

set.seed(427)
```

# Setting Up Multi-Modal Data

We generate both modalities from one shared cell embedding, which is exactly
the model `runMultiModalNMF()` fits: a shared H across modalities, and a
separate W per modality. Six programs drive 400 genes and 250 peaks across
400 cells.

```{r prepare-data}
k_true <- 6
n_cells <- 400
n_genes <- 400
n_peaks <- 250

# one shared embedding, used by both modalities
H_shared <- matrix(runif(k_true * n_cells, 0, 1)^3, k_true, n_cells)
cell_program <- apply(H_shared, 2, which.max)

make_modality <- function(n_features, sd_noise) {
    W <- matrix(runif(n_features * k_true, 0, 0.2), n_features, k_true)
    block <- split(seq_len(n_features),
                   rep(seq_len(k_true), length.out = n_features))
    for (j in seq_len(k_true)) W[block[[j]], j] <- runif(length(block[[j]]), 1, 3)
    mat <- W %*% H_shared
    mat <- mat + abs(matrix(rnorm(length(mat), sd = sd_noise), nrow = n_features))
    log1p(mat / max(mat) * 100)
}

rna_mat <- make_modality(n_genes, 0.05)
rownames(rna_mat) <- sprintf("Gene_%04d", seq_len(n_genes))
colnames(rna_mat) <- sprintf("Cell_%03d", seq_len(n_cells))

atac_mat <- make_modality(n_peaks, 0.10)
rownames(atac_mat) <- sprintf("peak_%04d", seq_len(n_peaks))
colnames(atac_mat) <- colnames(rna_mat)

sce <- SingleCellExperiment(list(logcounts = rna_mat))
sce$program <- paste0("P", cell_program)
altExp(sce, "ATAC") <- SingleCellExperiment(list(counts = atac_mat))

sce
```

The SCE now has two modalities:

- **Main assay** ("logcounts"): 400 genes
- **altExp "ATAC"**: 250 peaks

Both share the same cells.

# Running Multi-Modal NMF

`runMultiModalNMF()` accepts modalities via the `alt_exps` parameter. Use `"main"` to refer to the primary SCE assay:

```{r run-multimodal}
sce <- runMultiModalNMF(sce, k = k_true,
                        alt_exps = c("main", "ATAC"),
                        main_assay = "logcounts",
                        alt_assay = "counts",
                        modality_names = c("RNA", "ATAC"),
                        seed = 1, verbose = FALSE)
```

# Shared Cell Embedding

The shared H matrix captures cell identity using information from both modalities:

```{r shared-embedding}
shared <- reducedDim(sce, "MultiNMF")
dim(shared)

# the shared embedding recovers the simulated programs
table(truth = sce$program,
      recovered = paste0("R", apply(shared, 1, which.max)))
```

For comparison, a single-modality fit on RNA alone:

```{r single-modality}
sce <- runNMFscape(sce, k = k_true, name = "NMF_RNA", seed = 1,
                   verbose = FALSE)
dim(reducedDim(sce, "NMF_RNA"))
```

# Per-Modality Feature Loadings

Each modality gets its own basis matrix. Use the `modality` parameter in `getBasis()`:

```{r modality-basis}
rna_basis <- getBasis(sce, "MultiNMF", modality = "RNA")
cat("RNA basis:", dim(rna_basis), "\n")

atac_basis <- getBasis(sce, "MultiNMF", modality = "ATAC")
cat("ATAC basis:", dim(atac_basis), "\n")
```

## Top features per modality

```{r top-features}
# Top RNA genes per program
top_rna <- getTopFeatures(sce, name = "MultiNMF", n = 5, modality = "RNA")
cat("Top RNA genes for Program 1:\n")
top_rna[[1]]

# Top peaks per program
top_atac <- getTopFeatures(sce, name = "MultiNMF", n = 5, modality = "ATAC")
cat("\nTop peaks for Program 1:\n")
top_atac[[1]]
```

## Program usage across cell types

The shared embedding can be visualized the same way as standard NMF:

```{r dotplot, fig.width=6, fig.height=5}
plotProgramDots(sce, group = "program", nmf_name = "MultiNMF")
```

# When to Use Multi-Modal NMF

Real use cases for `runMultiModalNMF()`:

- **10x Multiome**: Joint RNA + ATAC-seq from the same nuclei. Store ATAC peaks as `altExp`.
- **CITE-seq**: RNA + surface protein (ADT). Protein panel as `altExp`.
- **SHARE-seq**: RNA + ATAC with combinatorial indexing.
- **Spatial + scRNA-seq**: When cells are matched across modalities.

The key advantage over running NMF separately on each modality is that the shared embedding enforces consistency --- programs must explain variation in *both* data types, filtering out modality-specific technical noise.

Note that multi-modal results are stored as a `factor_net_result`, so
`getBasis()` requires a `modality` and `predictNMF()` does not apply; see
`?getBasis` and `?predictNMF`.

**Note:** This vignette used simulated data for illustration. With real multiome data, the multi-modal embedding typically outperforms single-modality NMF for cell type separation and reveals regulatory relationships between gene expression and chromatin state.

# Session Information

```{r session-info}
sessionInfo()
```
