---
title: "NMFscape: Package Overview"
author: "Michael Totty"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{NMFscape: Package Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

# Overview

NMFscape provides a Bioconductor integration layer for [RcppML](https://github.com/zdebruine/RcppML), enabling fast non-negative matrix factorization (NMF) on `SingleCellExperiment` and `SpatialExperiment` objects. It wraps RcppML's computation engine and adds SCE-native storage, visualization, and biological interpretation tools.

# Function Reference

## Core NMF

| Function | Description |
|---|---|
| `runNMFscape()` | Standard NMF with 6 loss distributions |
| `runDeepNMF()` | Hierarchical multi-layer NMF (FactorNet) |
| `runMultiModalNMF()` | Joint NMF across modalities (RNA + ATAC) |
| `runConditionedNMF()` | Batch-conditioned NMF |
| `consensusNMF()` | Consensus NMF with stability metrics |
| `runFactorNet()` | Store custom FactorNet results in SCE |

## Diagnostics and Model Selection

| Function | Description |
|---|---|
| `selectRank()` | Cross-validation for optimal k |
| `plotRankSelection()` | Visualize CV results |
| `selectDistribution()` | Compare loss distributions (MSE, Poisson, NB) |
| `evaluateNMF()` | Reconstruction loss for a fitted model |
| `assessNMF()` | Embedding quality vs known labels (ARI, NMI) |

## Transfer Learning and Refinement

| Function | Description |
|---|---|
| `predictNMF()` | Project new data onto an existing model |
| `refineNMF()` | Label-guided model refinement |

## Accessors

| Function | Description |
|---|---|
| `getBasis()` | Gene loadings matrix (W), with modality support |
| `getCoefficients()` | Cell embeddings matrix (H) |
| `getTopFeatures()` | Top genes per program |
| `reconstructNMF()` | Reconstruct the original matrix |
| `getModel()` | Raw RcppML S4 model object |
| `getDiagonal()` | Diagonal scaling vector (d) |

## Visualization

| Function | Description |
|---|---|
| `vizDimRed()` | NMF programs on any dimension reduction |
| `vizUMAP()` | NMF programs on UMAP |
| `plotProgramDots()` | Dot plot of program usage by group |
| `plotProgramHeatmap()` | Heatmap of top genes per program |
| `plotRankSelection()` | Cross-validation elbow plot |

## Differential Program Analysis

| Function | Description |
|---|---|
| `FindAllDEPs()` | 1-vs-all differential program enrichment |
| `plotDEPsVolcano()` | Volcano plots of enrichment results |
| `plotDEPsHeatmap()` | Heatmap of fold enrichment across cell types |

# Vignettes

- **Getting Started**: Basic NMF workflow on mouse brain data. `vignette("getting-started")`
- **Choosing k**: Rank selection via cross-validation, distribution selection, and consensus stability. `vignette("choosing-k")`
- **Cell Type Programs**: Biological interpretation and label-guided refinement. `vignette("cell-type-programs")`
- **Deep NMF**: Hierarchical gene programs via multi-layer factorization. `vignette("deep-nmf")`
- **Multi-Modal NMF**: Joint RNA + chromatin factorization. `vignette("multi-modal")`
- **Transfer Learning**: Reference mapping and batch correction. `vignette("transfer-learning")`

# Quick Start

```{r quickstart, eval = FALSE}
library(NMFscape)
library(scuttle)
library(scRNAseq)

sce <- ZeiselBrainData()
sce <- sce[!grepl("^mt-", rownames(sce)), ]
sce <- logNormCounts(sce)

# Run NMF
sce <- runNMFscape(sce, k = 10)

# Explore
plotProgramDots(sce, group = "level1class")
plotProgramHeatmap(sce, n_genes = 10)
getTopFeatures(sce, n = 5)
```

# Session Information

```{r session-info}
sessionInfo()
```
