---
title: "Spatially-Aware NMF"
author: "Michael Totty"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Spatially-Aware NMF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3,
  fig.retina = 1,
  dpi = 96
)
```

# Introduction

Plain NMF treats every spot in a spatial experiment as an independent sample.
It never sees the tissue. Programs that are genuinely regional therefore come
out speckled: neighbouring spots that should share a program disagree simply
because each one carries only a handful of counts.

`runSpatialNMF()` adds a graph-Laplacian smoothness penalty to the objective:

$$\|A - WH\|_F^2 + \lambda \cdot \mathrm{tr}(H L H^\top)$$

where $L$ is the Laplacian of a spatial neighbour graph. That trace equals
$\tfrac{1}{2}\sum_{ij} A_{ij}\|h_i - h_j\|^2$, so the penalty is exactly the
total disagreement in program usage between neighbouring spots. Turning
$\lambda$ up buys spatial coherence at the cost of blurring real boundaries;
this vignette shows both ends of that trade.

```{r load, message = FALSE}
library(NMFscape)
library(SpatialExperiment)
library(ggplot2)
```

# A synthetic tissue

We simulate a 20x20 grid of spots split into four contiguous quadrant
domains. Each domain is driven by its own 30-gene program on top of a shared
background. Counts are Poisson and deliberately shallow --- a mean well below
one count per gene per spot --- which is the regime real Visium data lives in
and the regime where a single spot tells you almost nothing on its own.

```{r simulate}
set.seed(1)
side <- 20
n_genes <- 120

grid <- expand.grid(x = seq_len(side), y = seq_len(side))
half <- side / 2
domain <- ifelse(grid$x <= half & grid$y <= half, 1L,
          ifelse(grid$x >  half & grid$y <= half, 2L,
          ifelse(grid$x <= half & grid$y >  half, 3L, 4L)))

n_spots <- nrow(grid)
mu <- matrix(0.3, n_genes, n_spots)
for (d in seq_len(4)) {
    mu[((d - 1) * 30 + 1):(d * 30), domain == d] <- 0.6
}

counts <- matrix(rpois(n_genes * n_spots, mu), n_genes, n_spots)
rownames(counts) <- paste0("gene", seq_len(n_genes))
colnames(counts) <- paste0("spot", seq_len(n_spots))

spe <- SpatialExperiment(
    assays = list(logcounts = log1p(counts)),
    spatialCoords = as.matrix(grid)
)
spe$domain <- factor(domain)

mean(counts)
```

A helper to plot program scores in tissue space:

```{r plot-helper}
plotPrograms <- function(x, name, title) {
    scores <- reducedDim(x, name)
    coords <- spatialCoords(x)
    df <- do.call(rbind, lapply(seq_len(ncol(scores)), function(i) {
        data.frame(
            x = coords[, 1], y = coords[, 2],
            score = scores[, i] / max(scores[, i]),
            program = colnames(scores)[i]
        )
    }))
    ggplot(df, aes(x = x, y = y, fill = score)) +
        geom_raster() +
        facet_wrap(~ program, nrow = 1) +
        scale_fill_viridis_c(guide = "none") +
        coord_fixed() +
        theme_void() +
        ggtitle(title)
}
```

# Plain NMF is speckled

```{r plain}
plain <- runNMFscape(spe, k = 4, seed = 7, verbose = FALSE)
plotPrograms(plain, "NMF", "runNMFscape(): salt-and-pepper")
```

The four quadrants are visible, but each is grainy: adjacent spots inside one
domain disagree sharply about how much of the program they use.

`spatialAutocorrelation()` puts a number on that. Moran's I is near
$-1/(n-1) \approx 0$ for a spatially random pattern and approaches 1 for
contiguous domains.

```{r plain-moran}
spatialAutocorrelation(plain, name = "NMF")
```

# Spatially-aware NMF is coherent

```{r spatial}
spatial <- runSpatialNMF(spe, k = 4, graph = "knn", n_neighbors = 6,
                         graph_lambda = 1, seed = 7, verbose = FALSE)
plotPrograms(spatial, "SpatialNMF", "runSpatialNMF(): coherent domains")
```

```{r spatial-moran}
spatialAutocorrelation(spatial)
```

Moran's I rises from roughly 0.5 to roughly 0.9. Agreement with the true
domains improves too --- the penalty is not just cosmetic smoothing:

```{r ari}
adjustedRandIndex <- function(a, b) {
    tab <- table(a, b)
    choose2 <- function(z) z * (z - 1) / 2
    expected <- sum(choose2(rowSums(tab))) * sum(choose2(colSums(tab))) /
        choose2(sum(tab))
    (sum(choose2(tab)) - expected) /
        ((sum(choose2(rowSums(tab))) + sum(choose2(colSums(tab)))) / 2 -
             expected)
}

c(
    plain = adjustedRandIndex(max.col(reducedDim(plain, "NMF")), spe$domain),
    spatial = adjustedRandIndex(max.col(reducedDim(spatial, "SpatialNMF")),
                                spe$domain)
)
```

# The lambda dial

`graph_lambda` is dimensionless: it is rescaled internally against the
unregularized fit, so the same value means the same strength of smoothing on
any dataset. Sweeping it shows the bias-variance trade-off directly.

```{r sweep, fig.height = 3.2}
lambdas <- c(0, 0.03, 0.1, 0.3, 1, 3, 10)

sweep_result <- do.call(rbind, lapply(lambdas, function(lam) {
    fit <- runSpatialNMF(spe, k = 4, graph_lambda = lam, seed = 7,
                         verbose = FALSE)
    data.frame(
        lambda = lam,
        moran_I = mean(spatialAutocorrelation(fit)$moran_I),
        ARI = adjustedRandIndex(max.col(reducedDim(fit, "SpatialNMF")),
                                spe$domain)
    )
}))

sweep_result
```

```{r sweep-plot, fig.height = 3}
long <- rbind(
    data.frame(lambda = sweep_result$lambda, value = sweep_result$moran_I,
               metric = "Moran's I (spatial coherence)"),
    data.frame(lambda = sweep_result$lambda, value = sweep_result$ARI,
               metric = "ARI vs true domains (accuracy)")
)

ggplot(long, aes(x = lambda + 0.01, y = value)) +
    geom_line() +
    geom_point() +
    facet_wrap(~ metric) +
    scale_x_log10() +
    labs(x = "graph_lambda (log scale, offset for lambda = 0)", y = NULL) +
    theme_bw()
```

Moran's I climbs monotonically --- you can always buy more smoothness. ARI
peaks and then falls: past the optimum the penalty starts erasing the real
boundaries between domains along with the noise. **Moran's I alone cannot
choose lambda for you.** It tells you when smoothing has stopped adding
coherence; it will happily keep rising while accuracy collapses.

# Choosing the graph and lambda

**The graph.** `"knn"` with `n_neighbors = 6` is the default and matches the
Visium lattice, where each spot has six immediate neighbours. Use
`"distance"` when spot density varies and you want a fixed physical
neighbourhood rather than a fixed count; without an explicit `radius` it picks
the radius that gives the median spot about `n_neighbors` neighbours.
`"delaunay"` (requires `deldir`) adapts to irregular layouts and never leaves
a spot isolated, which makes it a good fit for single-cell resolution
platforms and for tissue with holes.

Keep `normalize_laplacian = TRUE` (the default) unless you have a reason not
to. Spots at a tissue edge have fewer neighbours, and the combinatorial
Laplacian penalizes them proportionally less, so edges stay noisier than the
interior.

**Lambda.** Start at the default of 1 and sweep an order of magnitude either
side. Two signals to combine:

- Moran's I, from `spatialAutocorrelation()`, should rise steeply and then
  flatten. The knee is where extra smoothing stops buying coherence.
- Reconstruction error, from `evaluateNMF()`, should stay close to the
  unregularized fit. A sharp rise means the penalty is now fighting the data.

Take the largest lambda before reconstruction error takes off. If you have
any ground truth --- annotated layers, a histology segmentation, marker genes
with known distributions --- score against it directly, as this vignette does
with ARI, and trust that over either proxy.

**Beyond coordinates.** Passing a matrix to `graph` instead of a method name
uses your own spot-by-spot adjacency, which is how you run this on a plain
`SingleCellExperiment`, stitch several tissue sections into one graph, or use
a histology-aware neighbourhood. `feature_graph` and `feature_lambda` apply
the same machinery to the gene loadings, for a co-expression or
protein-protein interaction prior.

# A note on the implementation

RcppML 1.0.0 documents `graph_W`, `graph_H` and `graph_lambda` arguments on
`nmf()`. `runSpatialNMF()` does not use them, and rejects them if passed
through `...`. In testing they change the fit without making the factors
smoother over the supplied graph --- a randomly rewired graph with the same
degree sequence perturbs the result as much as the true one, and relative
smoothness over the graph's own edges gets *worse* as the penalty rises. What
looks like smoothing there is uniform shrinkage of H.

Instead, `RcppML::nmf()` provides the initialization and the penalty is
applied by graph-regularized multiplicative updates (Cai et al. 2011), which
split each Laplacian into its non-negative parts $L = D - S$ so that both
factors stay non-negative and the objective is non-increasing.

```{r session-info}
sessionInfo()
```
