---
title: "Transfer Learning and Batch Correction"
author: "Michael Totty"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Transfer Learning and Batch Correction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  fig.retina = 1,
  dpi = 96
)
```

# Introduction

A common workflow in single-cell genomics is to train a reference model on a well-annotated dataset and project new (query) data onto it. NMFscape supports this via `predictNMF()` for reference mapping and `runConditionedNMF()` for batch-corrected factorization when combining datasets.

This vignette demonstrates both approaches on two simulated datasets that
share cell types but differ in composition and technical offset --- the
situation reference mapping is meant to handle. Everything shown works
unchanged on real data; see the *Discovering Cell Type Programs* vignette for
a worked example on brain snRNA-seq.

```{r load-packages, message=FALSE}
library(NMFscape)
library(SingleCellExperiment)
library(scuttle)
library(ggplot2)
library(patchwork)

set.seed(427)
```

# Prepare Reference and Query

Both datasets are built from the same six cell-type programs over the same
gene set, but with different cell-type proportions and a batch-specific gain,
so they overlap biologically while differing technically.

```{r simulate}
k_true <- 6
n_genes <- 400
gene_ids <- sprintf("Gene_%04d", seq_len(n_genes))

# shared program definitions: each program drives a block of genes
W_true <- matrix(runif(n_genes * k_true, 0, 0.2), n_genes, k_true)
blocks <- split(seq_len(n_genes), rep(seq_len(k_true), length.out = n_genes))
for (j in seq_len(k_true)) {
    W_true[blocks[[j]], j] <- runif(length(blocks[[j]]), 1, 3)
}

make_dataset <- function(n_cells, proportions, batch_gain, prefix) {
    types <- sample(seq_len(k_true), n_cells, replace = TRUE,
                    prob = proportions)
    H <- matrix(runif(k_true * n_cells, 0, 0.15), k_true, n_cells)
    H[cbind(types, seq_len(n_cells))] <- runif(n_cells, 1, 3)

    mat <- (W_true * batch_gain) %*% H
    mat <- mat * exp(rnorm(length(mat), 0, 0.15))
    mat <- log1p(mat / max(mat) * 100)
    dimnames(mat) <- list(gene_ids, sprintf("%s_%03d", prefix,
                                            seq_len(n_cells)))

    sce <- SingleCellExperiment(list(logcounts = mat))
    sce$celltype <- paste0("Type_", types)
    sce
}

# reference: all six types, roughly balanced
ref <- make_dataset(500, rep(1, k_true), 1.0, "ref")

# query: same types, skewed composition and a different technical gain
query <- make_dataset(300, c(3, 3, 2, 1, 1, 1), 1.6, "qry")

ref
query
```

# Build Reference NMF Model

Train NMF on the reference dataset:

```{r train-reference}
ref <- runNMFscape(ref, k = k_true, seed = 1, verbose = FALSE)
```

```{r show-reference, fig.width=6, fig.height=5}
plotProgramDots(ref, group = "celltype") +
    ggtitle("Reference: Program Usage")
```

# Project Query Data

`predictNMF()` projects the query dataset onto the reference NMF model. It automatically aligns features (genes) between the two datasets, using only shared genes:

```{r project-query}
query <- predictNMF(query, reference = ref, verbose = FALSE)
```

The projected coefficients are stored in `reducedDim(query, "NMF_projected")`:

```{r show-projection}
dim(reducedDim(query, "NMF_projected"))
```

```{r projected-dotplot, fig.width=6, fig.height=5}
plotProgramDots(query, group = "celltype", nmf_name = "NMF_projected") +
    ggtitle("Query: Projected onto Reference Model")
```

Because the two datasets share program definitions, each query cell type
should light up the same reference program its counterpart does:

```{r projection-recovery}
projected <- reducedDim(query, "NMF_projected")
table(truth = query$celltype,
      program = paste0("NMF_", apply(projected, 1, which.max)))
```

## Compare to de novo NMF

We can also run NMF directly on the query and compare:

```{r denovo}
query <- runNMFscape(query, k = k_true, name = "NMF_denovo", seed = 1,
                     verbose = FALSE)
```

```{r compare-approaches, fig.width=14, fig.height=5}
p1 <- plotProgramDots(query, group = "celltype",
                      nmf_name = "NMF_projected") +
    ggtitle("Projected from Reference")

p2 <- plotProgramDots(query, group = "celltype",
                      nmf_name = "NMF_denovo") +
    ggtitle("De Novo on Query")

p1 + p2
```

Projection maps the query into the same program space as the reference, making the programs directly comparable across datasets. De novo NMF finds dataset-specific programs.

# Batch-Corrected NMF

When combining datasets for joint analysis, batch effects can dominate the factorization. `runConditionedNMF()` conditions out a batch variable during NMF so that programs capture biology rather than technical variation.

```{r combine-datasets}
shared_genes <- intersect(rownames(ref), rownames(query))

ref_sub <- ref[shared_genes, ]
query_sub <- query[shared_genes, ]

ref_sub$dataset <- "reference"
query_sub$dataset <- "query"

# keep only the shared colData columns before binding
keep <- c("celltype", "dataset")
colData(ref_sub) <- colData(ref_sub)[, keep]
colData(query_sub) <- colData(query_sub)[, keep]
reducedDims(ref_sub) <- list()
reducedDims(query_sub) <- list()
metadata(ref_sub) <- list()
metadata(query_sub) <- list()

combined <- cbind(ref_sub, query_sub)
combined
```

```{r conditioned-nmf}
# Standard NMF (no batch correction)
combined <- runNMFscape(combined, k = k_true, name = "NMF_uncorrected",
                        seed = 1, verbose = FALSE)

# Batch-conditioned NMF
combined <- runConditionedNMF(combined, k = k_true,
                              condition_col = "dataset",
                              seed = 1, verbose = FALSE)
```

```{r compare-batch, fig.width=14, fig.height=5}
p1 <- plotProgramDots(combined, group = "dataset",
                      nmf_name = "NMF_uncorrected") +
    ggtitle("Uncorrected NMF")

p2 <- plotProgramDots(combined, group = "dataset",
                      nmf_name = "CondNMF") +
    ggtitle("Batch-Conditioned NMF")

p1 + p2
```

In uncorrected NMF, some programs may reflect dataset identity rather than cell type. Conditioned NMF should show more balanced program usage across datasets for shared cell types.

# Practical Guidance

**When projection works well:**

- Reference and query share major cell types
- The reference is comprehensive and well-annotated
- You want programs that are directly comparable across datasets

**When projection may fail:**

- The query contains cell types absent from the reference
- Technical differences (protocol, sequencing depth) are extreme
- Gene detection patterns differ substantially

**When to use conditioned NMF instead:**

- You want a joint embedding of multiple datasets
- Both datasets contribute unique cell types
- You need to explicitly model and remove batch effects

For large-scale atlas integration, consider running `runConditionedNMF()` on the combined data, then using `assessNMF()` to verify that cell types cluster by biology rather than dataset.

# Session Information

```{r session-info}
sessionInfo()
```
