---
title: "Transfer Learning and Batch Correction"
author: "Michael Totty"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Transfer Learning and Batch Correction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 6,
  eval = FALSE
)
```

*Note: This vignette contains code that downloads datasets and runs
computationally intensive analyses. Set `eval = TRUE` or run chunks
interactively to execute.*

# Introduction

A common workflow in single-cell genomics is to train a reference model on a well-annotated dataset and project new (query) data onto it. NMFscape supports this via `predictNMF()` for reference mapping and `runConditionedNMF()` for batch-corrected factorization when combining datasets.

This vignette demonstrates both approaches using two mouse brain datasets: the Zeisel cortex/hippocampus atlas as the reference and the Tasic visual cortex dataset as the query.

```{r load-packages, message=FALSE}
library(NMFscape)
library(SingleCellExperiment)
library(scuttle)
library(scater)
library(scRNAseq)
library(ggplot2)
library(patchwork)

set.seed(427)
```

# Prepare Reference and Query

```{r prepare-reference}
# Reference: Zeisel brain (~3k cells, broad cell types)
ref <- ZeiselBrainData()
ref <- ref[!grepl("^mt-", rownames(ref)), ]
ref <- logNormCounts(ref)
ref <- runPCA(ref, ncomponents = 20)
ref <- runUMAP(ref, dimred = "PCA")
```

```{r prepare-query}
# Query: Tasic visual cortex (~1.8k cells, detailed subtypes)
query <- TasicBrainData()
query <- query[, query$broad_type != "Unclassified"]
query <- logNormCounts(query)
query <- runPCA(query, ncomponents = 20)
query <- runUMAP(query, dimred = "PCA")
```

# Build Reference NMF Model

Train NMF on the reference dataset:

```{r train-reference}
ref <- runNMFscape(ref, k = 10, verbose = FALSE)
```

```{r show-reference, fig.width=6, fig.height=5}
plotProgramDots(ref, group = "level1class") +
    ggtitle("Reference: Zeisel Program Usage")
```

# Project Query Data

`predictNMF()` projects the query dataset onto the reference NMF model. It automatically aligns features (genes) between the two datasets, using only shared genes:

```{r project-query}
query <- predictNMF(query, reference = ref, verbose = FALSE)
```

The projected coefficients are stored in `reducedDim(query, "NMF_projected")`:

```{r show-projection}
dim(reducedDim(query, "NMF_projected"))
```

```{r projected-dotplot, fig.width=6, fig.height=5}
plotProgramDots(query, group = "broad_type", nmf_name = "NMF_projected") +
    ggtitle("Query: Tasic Projected onto Zeisel Model")
```

## Compare to de novo NMF

We can also run NMF directly on the query and compare:

```{r denovo}
query <- runNMFscape(query, k = 10, name = "NMF_denovo", verbose = FALSE)
```

```{r compare-approaches, fig.width=14, fig.height=5}
p1 <- plotProgramDots(query, group = "broad_type",
                      nmf_name = "NMF_projected") +
    ggtitle("Projected from Reference")

p2 <- plotProgramDots(query, group = "broad_type",
                      nmf_name = "NMF_denovo") +
    ggtitle("De Novo on Query")

p1 + p2
```

Projection maps the query into the same program space as the reference, making the programs directly comparable across datasets. De novo NMF finds dataset-specific programs.

# Batch-Corrected NMF

When combining datasets for joint analysis, batch effects can dominate the factorization. `runConditionedNMF()` conditions out a batch variable during NMF so that programs capture biology rather than technical variation.

```{r combine-datasets}
# Find shared genes
shared_genes <- intersect(rownames(ref), rownames(query))

# Subset and combine
ref_sub <- ref[shared_genes, ]
query_sub <- query[shared_genes, ]

# Add dataset labels
ref_sub$dataset <- "Zeisel"
query_sub$dataset <- "Tasic"

# Combine into one SCE
combined <- cbind(ref_sub, query_sub)
combined <- logNormCounts(combined)
```

```{r conditioned-nmf}
# Standard NMF (no batch correction)
combined <- runNMFscape(combined, k = 10, name = "NMF_uncorrected",
                        verbose = FALSE)

# Batch-conditioned NMF
combined <- runConditionedNMF(combined, k = 10, condition_col = "dataset",
                              verbose = FALSE)
```

```{r compare-batch, fig.width=14, fig.height=5}
p1 <- plotProgramDots(combined, group = "dataset",
                      nmf_name = "NMF_uncorrected") +
    ggtitle("Uncorrected NMF")

p2 <- plotProgramDots(combined, group = "dataset",
                      nmf_name = "CondNMF") +
    ggtitle("Batch-Conditioned NMF")

p1 + p2
```

In uncorrected NMF, some programs may reflect dataset identity rather than cell type. Conditioned NMF should show more balanced program usage across datasets for shared cell types.

# Practical Guidance

**When projection works well:**

- Reference and query share major cell types
- The reference is comprehensive and well-annotated
- You want programs that are directly comparable across datasets

**When projection may fail:**

- The query contains cell types absent from the reference
- Technical differences (protocol, sequencing depth) are extreme
- Gene detection patterns differ substantially

**When to use conditioned NMF instead:**

- You want a joint embedding of multiple datasets
- Both datasets contribute unique cell types
- You need to explicitly model and remove batch effects

For large-scale atlas integration, consider running `runConditionedNMF()` on the combined data, then using `assessNMF()` to verify that cell types cluster by biology rather than dataset.

# Session Information

```{r session-info, eval = TRUE}
sessionInfo()
```
